msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-05 paper.tex
Update main section text.

Kingman's coalescent assumes that at most two ancestral lineages can
merge at each merger event.  This follows from the fact that the
Kingman coalescent can be shown to be the process describing the
random ancestral relations of gene copies sampled from a population
evolving according to the Wright-Fisher model (or similar models) of
genetic evolution.  Multiple-merger coalescent models can also be
obtained from appropriate extensions of the Moran
model \citep{EW06,HM12}, i.e.\ where a single randomly picked individual
(e.g.\ in a haploid population), or a pair of diploid parent
individuals \citep{BBE13}, produces offspring in each reproduction
event.  Convergence to the Kingman coalescent follows from certain
assumptions on the offspring distribution \citep{S99,MS01,SW08}.  These
assumptions may be violated in certain highly fecund
organisms \citep{hedgecock_94,B94,HP11,A04,irwin16}.  We take `high
fecundity' to mean the ability of individuals to produce numbers of
offspring on the order of the population size.  A key issue regarding
high fecundity involves the concept of `sweepstakes
reproduction' \citep{hedgecock_94}, where a few individuals may produce
the bulk of the offspring in any given generation.  Sweepstakes
reproduction is not captured by the Wrigth-Fisher model (or similar
models).  Population models, which do capture high fecundity and
sweepstakes reproduction, are in the domain of attraction of
`multiple-merger' coalescents \citep{DK99,P99,S99,S00}, in which a
random number of ancestral lineages may merge at any given time.  The
term `multiple merger' refers to a merger of at least three ancestral
lineages, or the simultaneous \citep{S00,MS01} merger of at least four
lineages in at least two distinct groups, and each group involving at
least two lineages.  Multiple-merger coalescent processes have also
been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

\cite{becheler2020occupancy}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.

%assumption can be
%violated in some situations. Lambda coalescents
%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
%%HybridLambda~\citep{zhu2015hybrid}

>>>>

Kingman's coalescent assumes that at most two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

Although multiple merger coalescents have been of significant theoretical
interest for around two decades, there has been little practical software
available to simulate these models.
\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
simulate a related spatial continuum model~\citep{barton2010new},
and \cite{zhu2015hybrid} simulate genealogies within a species tree
based on a multiple-merger model.
More recently \cite{becheler2020occupancy}
provide a general method for simulating multiple merger processes
as part of the Quetzal framework~\citep{becheler2019quetzal}.
None of these methods work with large genomes, and very little work
has been performed on simulating multiple merger processes with recombination.

We have added two multiple merger coalescent models in \msprime\ 1.0, the
$\beta$-coalescent and ``dirac''-coalescent, allowing us to simulate
such models with recombination efficiently for the first time.
Please see the Appendix for full details and derivations.
